Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model

### 2024-01-04 analysis/paper/discussion.qmd
rename sections to .qmd

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 $\mu$m in
maximum length were underrepresented in the calculus samples compared to the original 
starch solutions, an effect that was consistent across all three treatments. Medium
granules (10--20 $\mu$m) were often over-represented
(Table \@ref(tab:sample-prop-tbl), 
and 
Figure \@ref(fig:ratio-plots)).
Large potato granules were most affected, potentially because of the greater
size-range. They can reach up to 100 $\mu$m in maximum length, whereas wheat 
granules generally only reach up to 35 $\mu$m 
[@gismondiStarchGranules2019; @haslamDecompositionStarch2004; @seidemannStarchAtlas1966, 174-176].
Granule morphology may also play a role. Large wheat granules 
are lenticular and have a larger surface area compared 
to volume, whereas large potato granules are ovoid and have a larger volume
compared to surface area 
[@vandeveldeStarchMorphology2002; @janeAnthologyStarch1994; @reichertStarchBible1913b, 364-365; @seidemannStarchAtlas1966, 174-176]. 
Another potentially important factor 
from our results is the size of the calculus deposit. We found a 
`r strength_cor` `r direct_cor` correlation between size of biofilm deposit and
retained starch granules (Figure \@ref(fig:cor-plot)A), meaning larger calculus
deposits contain a higher quantity of granules;
a result that contradicts findings from archaeological contexts
[@wesolowskiEvaluatingMicrofossilContent2010; @dudgeonDietGeography2014].
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \@ref(fig:cor-plot)B).
While the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches. This may also explain
the lower mean retention of starch granules in mixed treatments compared to
wheat treatments. Wheat treatment samples (mean = `r subset(anova_tbl, treatment == "wheat")$mean` mg) were on average larger than mixed treatment samples (mean = `r subset(anova_tbl, treatment == "mix")$mean` mg) (Table \@ref(tab:anova-tbl)); and while mixed treatment solutions contained the highest mean overall
granule counts, wheat treatment samples had the highest mean starch retention.
Further research is needed to determine why this differs from previous archaeological
findings.

@@

[@radiniFoodPathways2017; @radiniMedievalWomenEarly2019].  

>>>>

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 $\mu$m in
maximum length were underrepresented in the calculus samples compared to the original 
starch solutions, an effect that was consistent across all three treatments. Medium
granules (10--20 $\mu$m) were often over-represented
(Table \@ref(tab:sample-prop-tbl), 
and 
Figure \@ref(fig:ratio-plots)).
Large potato granules were most affected, potentially because of the greater
size-range. They can reach up to 100 $\mu$m in maximum length, whereas wheat 
granules generally only reach up to 35 $\mu$m 
[@gismondiStarchGranules2019; @haslamDecompositionStarch2004; @seidemannStarchAtlas1966, 174-176].
Granule morphology may also play a role. Large wheat granules 
are lenticular and have a larger surface area compared 
to volume, whereas large potato granules are ovoid and have a larger volume
compared to surface area 
[@vandeveldeStarchMorphology2002; @janeAnthologyStarch1994; @reichertStarchBible1913b, 364-365; @seidemannStarchAtlas1966, 174-176]. 
Another potentially important factor 
from our results is the size of the calculus deposit. We found a 
`r strength_cor` `r direct_cor` correlation between size of biofilm deposit and
retained starch granules (Figure \@ref(fig:cor-plot)A), meaning larger calculus
deposits contain a higher quantity of granules;
a result that contradicts findings from archaeological contexts
[@wesolowskiEvaluatingMicrofossil2010; @dudgeonDietGeography2014].
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \@ref(fig:cor-plot)B).
While the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches. This may also explain
the lower mean retention of starch granules in mixed treatments compared to
wheat treatments. Wheat treatment samples (mean = `r subset(anova_tbl, treatment == "wheat")$mean` mg) were on average larger than mixed treatment samples (mean = `r subset(anova_tbl, treatment == "mix")$mean` mg) (Table \@ref(tab:anova-tbl)); and while mixed treatment solutions contained the highest mean overall
granule counts, wheat treatment samples had the highest mean starch retention.
Further research is needed to determine why this differs from previous archaeological
findings.

@@

[@radiniFoodPathways2017; @radiniMedievalWomen2019].  

### 2024-01-04 analysis/paper/intro.qmd
rename sections to .qmd

<!-- basic introduction to the field, comprehensible to a scientist in any discipline -->
<!-- Dental calculus is mineralised plaque. And it is pretty cool. -->
Dental calculus has proven to contain a wealth of 
dietary information in the form of plant microfossils 
[@henryCalculusSyria2008; @hardyStarchGranulesDental2009], 
proteins [@hendyProteomicCalculus2018; @warinnerEvidenceMilk2014],
and other organic residues [@buckleyDentalCalculus2014]. 
This dietary information can be preserved within the mineralised dental plaque 
over many millennia, providing a unique window into the food-related behaviours of 
past populations 
[@henryCalculusSyria2008; @jovanovicNeolithicCalculus2021; @taoWheatCalculus2020] 
and extinct species [@henryNeanderthalCalculus2014; @hardyNeanderthalMedics2012].

@@

<!-- clearly stating the general problem being addressed by this particular study -->
Despite this, our knowledge of dental calculus and the incorporation pathways of 
the various markers is limited [@radiniFoodPathways2017], as is our 
knowledge of information-loss caused by these pathways. Additionally, the methods
we use to extract and analyse dental calculus, and make inferences on past diets
represent another potential source of bias. Studies on both archaeological and
modern individuals have explored these biases in more detail.
Extraction methods were tested 
by Tromp and colleagues [-@trompEDTACalculus2017], specifically regarding
decalcification using HCl or EDTA. 
The authors found significantly more starches with the EDTA extraction method 
than the HCl extraction method; however, as noted by the authors, comparisons 
involving archaeological calculus are problematic due to variability between and 
within individuals. 
Studies conducted on modern humans [@leonardDentalCalculus2015] 
and non-human primates [@powerChimpCalculus2015; @powerRepresentativenessDental2021] 
have explored how well microremains (phytoliths and starches)
extracted from dental calculus represent the actual dietary intake.
These studies are justifiably limited, 
despite meticulous documentation and observation, due to unknown variables and 
uncertainty involved in this kind of *in vivo* research. Dental calculus is a complex 
oral biofilm with a multifactorial aetiology and variable formation rates both 
within and between individuals [@jepsenCalculusRemoval2011; @haffajeeBiofilmPosition2009], 
contributing to
the stochasticity of starch representation being observed in numerous studies.
Additionally, the concentration of oral $\alpha$-amylase differs both between and
within individuals [@froehlichEffectOral1987; @naterHumanAmylase2005],
causing different rates of hydrolysis of the starch granules present in the oral
cavity. Add to this the effects of the many different methods
of starch processing [@hardyRecoveringInformation2018], as well as post-depositional
processes that are still being explored
[@graneroStarchTaphonomy2020; @mercaderExaggeratedExpectations2018],
and it becomes clear that using 
dental calculus to reconstruct diet is a highly unpredictable process.

>>>>

<!-- basic introduction to the field, comprehensible to a scientist in any discipline -->
<!-- Dental calculus is mineralised plaque. And it is pretty cool. -->
Dental calculus has proven to contain a wealth of 
dietary information in the form of plant microfossils 
[@henryCalculusSyria2008; @hardyStarchGranules2009], 
proteins [@hendyProteomicCalculus2018; @warinnerEvidenceMilk2014],
and other organic residues [@buckleyDentalCalculus2014]. 
This dietary information can be preserved within the mineralised dental plaque 
over many millennia, providing a unique window into the food-related behaviours of 
past populations 
[@henryCalculusSyria2008; @jovanovicNeolithicCalculus2021; @taoWheatCalculus2020] 
and extinct species [@henryNeanderthalCalculus2014; @hardyNeanderthalMedics2012].

@@

<!-- clearly stating the general problem being addressed by this particular study -->
Despite this, our knowledge of dental calculus and the incorporation pathways of 
the various markers is limited [@radiniFoodPathways2017], as is our 
knowledge of information-loss caused by these pathways. Additionally, the methods
we use to extract and analyse dental calculus, and make inferences on past diets
represent another potential source of bias. Studies on both archaeological and
modern individuals have explored these biases in more detail.
Extraction methods were tested 
by Tromp and colleagues [-@trompEDTACalculus2017], specifically regarding
decalcification using HCl or EDTA. 
The authors found significantly more starches with the EDTA extraction method 
than the HCl extraction method; however, as noted by the authors, comparisons 
involving archaeological calculus are problematic due to variability between and 
within individuals. 
Studies conducted on modern humans [@leonardPlantMicroremains2015] 
and non-human primates [@powerChimpCalculus2015; @powerRepresentativenessDental2021] 
have explored how well microremains (phytoliths and starches)
extracted from dental calculus represent the actual dietary intake.
These studies are justifiably limited, 
despite meticulous documentation and observation, due to unknown variables and 
uncertainty involved in this kind of *in vivo* research. Dental calculus is a complex 
oral biofilm with a multifactorial aetiology and variable formation rates both 
within and between individuals [@jepsenCalculusRemoval2011; @haffajeeBiofilmPosition2009], 
contributing to
the stochasticity of starch representation being observed in numerous studies.
Additionally, the concentration of oral $\alpha$-amylase differs both between and
within individuals [@froehlichEffectOral1987; @naterHumanAmylase2005],
causing different rates of hydrolysis of the starch granules present in the oral
cavity. Add to this the effects of the many different methods
of starch processing [@hardyRecoveringInformation2018], as well as post-depositional
processes that are still being explored
[@graneroStarchTaphonomy2020; @mercaderExaggeratedExpectations2018],
and it becomes clear that using 
dental calculus to reconstruct diet is a highly unpredictable process.

### 2024-01-04 analysis/paper/results.qmd
rename sections to .qmd

## No amylase activity detected in the model

@@

## Treatment type had minimal effect on biofilm growth

@@

## Starch counts

@@

### Proportion of available starches incorporated in samples

@@

### Size ratios differ between solutions and samples

@@

### Biofilm weight correlated positively with extracted starch counts

>>>>

### No amylase activity detected in the model

@@

### Treatment type had minimal effect on biofilm growth

@@

### Starch counts

@@

#### Proportion of available starches incorporated in samples

@@

#### Size ratios differ between solutions and samples

@@

#### Biofilm weight correlated positively with extracted starch counts
